SOURCE SEPARATION METHODS FOR COMPUTER-ASSISTED ORCHESTRATION

### 2022-04-20 source_separation/paper/MLSP/paper.tex
Move old paper to MLSP folder

\renewcommand{\baselinestretch}{0.96} % use the smallest value to make our paper 6 pages long (including references!)

@@

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
computer-assisted orchestration, sound source separation, Orchidea
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning.

	Orchestration~\cite{Adler2016} is hard to define, however one could say that orchestration is the set of musical writing techniques that focus on timbre, or \textit{tone color} of the sound. A simplistic  definition of musical orchestration would be to say that it consists of the distribution of the different musical voices of a piece to the instruments of the orchestra. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes.

	 Among the different approaches and techniques of computer-assisted orchestration, one of them is \textit{target-based orchestration}~\cite{Maresz2003}. Target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. This target corresponds to an audio extract, of the composer's choice, which can be musical or not. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints.

  Static orchestration is a type of orchestration that considers the target as a single vector of timbral features and creates only a single onset of note(s) as the solution, no matter if the target contains multiple onsets of sounds. In our experiments, we only perform static orchestrations.

@@

  We think source separation could be an important addition to improve orchestration as it allows for the orchestration of more complex sounds in a way that makes sense in an orchestral context. Orchestral music often uses multiple layers, so by separating the layers of a target and individually orchestrating them, the resulting orchestration may improve.

  For example, consider a target that has a continuous drone sound throughout and a melody playing above this drone. Without source separation, Orchidea would detect each note of the melody as a new onset and cut the target in time at each new note. However, this segmentation in time would also affect the drone, cutting it in time and forcing it to have a new onset each time the melody changes notes. The resulting orchestration would have multiple onsets for the drone, even though only one onset is needed at the beginning of the drone. When source separation is applied to this target, the drone and melody would be orchestrated separately. Therefore, the drone could be orchestrated by a single onset that lasts for the duration of the drone. At the same time, the melody could have as many new onsets as needed without affecting the drone. The orchestration would be greatly improved by removing the unnecessary onsets in the orchestration of the drone.

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and a selection of targets and orchestrations are available for listening at: \url{anonymous}.

@@

  We compare the effectiveness of different source separation methods for the task of orchestration by applying the separation methods to various targets, orchestrating the separated output of the method, and finally comparing this to the orchestration of the target if no separation was performed (see Fig. \ref{fig:full_diagram}). Following the previous example, consider a target sound that is a combination of a low droning sound and high-pitched whistle. With separation, these two sounds would be disentangled into two separate sub-targets and each would be individually orchestrated. The separated orchestrations would then be combined to create the solution. This would then be compared to the orchestration when no separation had been performed.

@@

    \subsection{Testing}\label{subsec:testing}
    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, \textit{without any separation performed}, is orchestrated using the entire orchestra. This creates what we call the \textit{full target orchestration}.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets. 
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textit{separated orchestrations}.
      \item The distances between target and orchestration are calculated for the full target orchestration and the separated orchestration.
    \end{enumerate}

@@

    \begin{figure}
    \centering
      \includegraphics[width=0.7\columnwidth]{figures/orch.jpg}
      \caption{Example waveforms showing onsets of the four sources in the target, and possible onsets in the full and separated orchestrations. In this example, the separation method could not distinguish between sources 2 and 3 and identified them as a single sub-target. Therefore, the separated orchestration has only 3 onsets, one of them being an orchestration that combines sources 2 and 3. Note that this is a simplified representation, in the actual targets sources overlap in time and are not as discrete as this diagram shows.}\label{fig:orchestrations}
    \end{figure}

    As described in Section \ref{subsec:data}, each target is comprised of four sources that are offset in time to begin playing at different times. Figure~\ref{fig:orchestrations} illustrates the effect this offset has on the full target and separated orchestrations. In the full target orchestration, all onsets occur at the beginning of the orchestration. Since there is no knowledge of different sources, Orchidea cannot place the different orchestrations at the appropriate offsets.

    For the separated orchestrations, the onsets of the sub-target orchestrations can be placed at the correct offset if the separation method has correctly separated out the sources. For this reason, despite only performing static orchestrations, the resulting separated orchestrations are offset in time compared to the full target orchestration.

@@

  \begin{table}
    \begin{center}
      \begin{tabular}{|c|c|}
        \hline
        & Average distance \\
        \hline
        Demucs & 25.11 \\
        \hline
        Open-Unmix & 24.48\\
        \hline
        TDCN & 26.44\\
        \hline
        \textbf{NMF} & \textbf{22.44} \\
        \hline
        TDCN++ & 24.38 \\
        \hline
        Full target orchestration & 24.96 \\
        \hline
      \end{tabular}
    \end{center}
    \caption{Average across 300 targets showing the distance between target and orchestration for various methods. For Demucs, Open-Unmix, TDCN, NMF, and TDCN++, the value shown is the average distance between the target sound and the ``separated orchestration'' created from the output of the given separation method (see Sec. \ref{subsec:testing}). Similarly, ``Full target orchestration'', as defined in Sec. \ref{subsec:testing}, is the average values between target and orchestrated solution. Lower values are better.}
    \label{tab:distances}
  \end{table}

@@

    \subsection{Results}
    The average distances for full target and separated orchestrations are displayed in Table \ref{tab:distances}. These results are the average distance between target and orchestration for 300 different targets. We found that NMF performed the best out of all separation methods. TDCN++, Open-Unmix, and NMF performed better than the full target orchestrations, meaning that separating targets using these methods leads to a better orchestration than if no separation is applied. However, Demucs and TDCN performed worse.

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. We believe this shows that source separation is a useful pre-processing stage for computer-assisted orchestration.

  In order to fully understand the impact of source separation on the orchestration process, we also conducted a qualitative evaluation of the generated orchestrations by acoustic inspection.
This evaluation showed that source separation significantly improves the quality or orchestrations from a musical standpoint. The next paragraphs will try to explain this in more detail.

	A major disadvantage of the orchestrations generated without any source separation is that, when two or more sources overlap in time, the orchestration algorithm cannot distinguish between them. As such, the orchestration appears to be more rigid: the whole orchestra suddenly changes configuration when a new source dominates, without any blending with the previous dominating source. This is against a well know orchestration principle called \emph{dovetailing} \cite[pp. 467-473]{Adler2016}. Dovetailing is a well known practice aimed at creating a blend of the instruments used during orchestration, in order to transition between timbres. Technically, dovetailing is the swapping and overlapping of musical lines between different instruments: musical lines are scattered across multiple instruments, connected with overlapping pitches.

	An interesting result that we discovered after acoustic inspection is that the orchestrations generated after source separation tend to favor dovetailing. Since only a portion of the orchestra is assigned to each source and since the sources are inherently asynchronous with each other, the final orchestration is more fluid and appears more interesting, from a musical standpoint.

  Quantitatively, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds. Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on an appropriate dataset of arbitrary sounds could yield better overall performance for orchestration.

>>>>

\renewcommand{\baselinestretch}{0.97} % Chercher la valeur la plus petite

@@

\begin{abstract}
  In this paper, we will study the possibility of adding source separation as a pre-processing step to the computer-assisted orchestration process. We first discuss the motivation of this addition and its potential to increase the quality of orchestrations of multi-layered sounds. Second, we select several state-of-the-art models for both music source separation (separation of instruments) and universal sound separation (separation of arbitrary sounds of different types), and compare their effectiveness for the task of orchestration. We assess which methods best suit the needs of orchestration by applying them on hand picked target sounds, orchestrating the separated outputs, and finally comparing them to the orchestration of the same target without separation. Our experiments show that the quality of orchestrations improves, both qualitatively and quantitatively, indicating that our approach is promising. Finally, we compare unsupervised methods to supervised methods for separation, and comment on the effect of training data selection on performance of supervised methods.
\end{abstract}
%
\begin{keywords}
computer-assisted orchestration, sound source separation, Orchidea
\end{keywords}
%
\section{Introduction}\label{sec:introduction}
  Computer-assisted composition is a field that focuses on the creation of computational tools to aid in the musical composition process \cite{FerVic2013, Ari2005}. Within this field, target-based computer-assisted musical orchestration is an example of how musical creativity can be supported by tools such as machine learning. Orchestration has been traditionally taught in an intuitive and non-formalized way, and computer-assisted orchestration seeks to help composers by accelerating certain creative processes. In particular, target-based computer-assisted orchestration helps composers explore new timbral possibilities by combining instrumental samples in such a way as to mimic the timbre of a given target sound. Formally, it is the process of creating an orchestral score that best matches an arbitrary target sound given a similarity metric and constraints~\cite{Maresz2003}. Static orchestration is a type of orchestration that considers the target as a single vector of timbral features and creates only a single onset of note(s) as the solution, no matter if the target contains multiple onsets of sounds. In our experiments, we only perform static orchestrations.

@@

%  Source separation is an important addition to improve orchestration as it allows for the orchestration of more complex sounds in a way that makes sense in an orchestral context. Orchestral music often uses multiple layers, so by separating the layers of a target and individually orchestrating them, the resulting orchestration may improve.

  For example, consider a target that has a continuous drone sound throughout and a melody playing above this drone. Without source separation, Orchidea would detect each note of the melody as a new onset and cut the target in time at each new note. However, this segmentation in time would also effect the drone, cutting it in time and forcing it to have a new onset each time the melody changes notes. The resulting orchestration would have multiple onsets for the drone, even though only one onset is needed at the beginning of the drone. When source separation is applied to this target, the drone and melody would be orchestrated separately. Therefore, the drone could be orchestrated by a single onset that lasts for the duration of the drone. At the same time, the melody could have as many new onsets as needed without affecting the drone. The orchestration would be greatly improved by removing the unnecessary onsets in the orchestration of the drone.

%  The codebase for this paper can be found at: \url{https://github.com/dzluke/MLSP2021}, and you can listen to a selection of targets and orchestrations here: \url{https://dzluke.github.io/MLSP2021/}.
  The codebase for this paper can be found at: \url{anonymous}, and you can listen to a selection of targets and orchestrations here: \url{anonymous}.

@@

  We compare the effectiveness of different source separation methods for the task of orchestration by applying the separation methods to various targets, orchestrating the separated output of the method, and finally comparing this to the orchestration of the target if no separation was performed (see Fig. \ref{fig:full_diagram}). Following the previous example, consider a target sound that is a combination of a low droning sound and high-pitched whistle. With separation, these two sounds would be disentangled into two separate sub-targets and each would be individually orchestrated. The separated orchestrations would then be combined to create the solution. This would then be compared to the orchestration if no separation had been performed.

@@

    \subsection{Testing}\label{subsec:testing}
    In order to compare the effectiveness of the separation methods, a full target orchestration and a ground truth orchestration are done for each target. The \textbf{full target orchestration} is the orchestration of the target without any separation performed. The \textbf{ground truth orchestration} takes the four sources that make up the target and orchestrates each one separately, then combines the orchestrations. This ground truth represents the orchestration that could take place if the separation method was perfect and separated a target exactly into its constituent sources.

    The procedure we use for testing is as follows:
    \begin{enumerate}
      \item The target is created as a combination of four randomly chosen sources, which are offset to begin playing at different times.
      \item The target, without any separation performed, is orchestrated using the entire orchestra. This is called the full target orchestration.
      \item Each separation method is performed on the target, each splitting the target into four sub-targets. 
      \item The full orchestra is randomly split into four equal sized subsets, each containing 7 to 8 instruments.
      \item For each method, the four sub-targets are separately orchestrated, each using a different subset of the orchestra, and then each orchestration is combined to play simultaneously. This creates the \textbf{separated orchestrations}.
      \item The ground truth orchestration is created by orchestrating the four sources, using the same orchestra subsets from step 4, and then combined to play simultaneously.
      \item The distances between target and orchestration are calculated for the full target orchestration, the separated orchestration, and the ground truth orchestration. 
    \end{enumerate}

@@

    \begin{figure}[t]
    \centering
      \includegraphics[width=0.7\columnwidth]{figures/orch.jpg}
      \caption{Example waveforms showing onsets of the four sources in the target, and possible onsets in the full, separated, and ground truth orchestrations. In this example, the separation method could not distinguish between sources 2 and 3 and identified them as a single sub-target. Therefore, the separated orchestration has only 3 onsets, one of them being an orchestration that combines sources 2 and 3. Note that this is a simplified representation, in the actual targets sources overlap in time and are not as discrete as this diagram shows.}\label{fig:orchestrations}
    \end{figure}

    As described in Section \ref{subsec:data}, each target is comprised of four sources that are offset in time to begin playing at different times. Figure~\ref{fig:orchestrations} illustrates the effect this offset has on the full target, separated, and ground truth orchestrations. In the full target orchestration, all onsets occur at the beginning of the orchestration. Since there is no knowledge of different sources, Orchidea cannot place the different orchestrations at the appropriate offsets.

    For the separated orchestrations, the onsets of the sub-target orchestrations can be placed at the correct offset if the separation method has correctly separated out the sources. In the ground truth orchestration, the sources are manually offset, leading to the ground truth having each source orchestration beginning at the correct time. For this reason, despite only performing static orchestrations, the resulting separated and ground truth orchestrations are offset in time compared to the full target orchestration.

@@

  \begin{table}[t]
    \begin{center}
      \begin{tabular}{|c|c|}
        \hline
        & Average distance \\
        \hline
        Demucs & 25.11 \\
        \hline
        Open-Unmix & 24.48\\
        \hline
        TDCN & 26.44\\
        \hline
        NMF & 22.44 \\
        \hline
        TDCN++ & 24.38 \\
        \hline
        Full target orchestration & 24.96 \\
        \hline
        Ground truth orchestration & 24.19 \\
        \hline
      \end{tabular}
    \end{center}
    \caption{Average across 300 targets showing the distance between target and orchestration for various methods. For Demucs, Open-Unmix, TDCN, NMF, and TDCN++, the value shown is the average distance between the target sound and the ``separated orchestration'' created from the output of the given separation method (see Sec. \ref{subsec:testing}). Similarly, ``Full target orchestration'' and ``Ground truth orchestration'', as defined in Sec. \ref{subsec:testing}, are the average values between target and orchestrated solution. Lower values are better.}
    \label{tab:distances}
  \end{table}

@@

    \subsection{Results}
    The average distances for full target, separated, and ground truth orchestrations are displayed in Table \ref{tab:distances}. These results are the average distance between target and orchestration for 300 different targets. We found that NMF performed the best out of all separation methods, and even outperformed the ground truth orchestrations, on average. TDCN++, Open-Unmix, and NMF performed better than the full target orchestrations, meaning that separating targets using these methods leads to a better orchestration than if no separation is applied.

  \section{Discussion}\label{sec:discussion}
  In this paper, we addressed the problem of using source separation as a pre-processing step for computer-assisted orchestration. The results show that performing source separation followed by a static orchestration of each individual sub-target improved the spectral distance between the original target and its orchestrated version in three out of five of the methods tested. This shows that source separation is a useful pre-processing stage for computer-assisted orchestration.

  Furthermore, the one unsupervised method we tested, NMF, outperformed all the supervised methods. This suggests that methods like NMF can still be useful for specific problems when available data is limited. It is logical that unsupervised methods will perform well for problems in which there is limited data, as they require less data than supervised methods. NMF also outperformed the ground truth orchestrations, which could be due to the relative simplicity of NMF's output subtargets making subsequent orchestration easier for Orchidea, but more experimentation would need to be done in order to confirm this. It is important to notice that for the supervised methods, we used pre-trained versions of the neural models. All of these models, except for TDCN++, were trained on musical data. However, we tested them on a wide range of targets, many of which were not strictly musical. This could explain why Demucs, Open-Unmix, and TDCN performed worse than NMF, and why TDCN++, which was trained on arbitrary sounds, outperformed the other supervised methods. Therefore, we may infer from our results that models trained on generic sounds generate a better separation for orchestration when compared to models trained only on musical sounds.

  Since unsupervised methods perform well, it is reasonable to think that supervised methods trained on a dataset of arbitrary sounds could yield better overall performance for orchestration.
